\LARGE{{\bf{Gravitational Wave Signals from Multiple Hidden Sectors} \\ }}

### 2019-10-02 gwnnDraft02.tex
Added standard sector neffective

\subsection{Standard Sectors}
As shown in Eq.~(\ref{eqn:vevs}), for standard sectors with increasing index $i$ the vevs of said sectors increase $v_i\propto \sqrt{i}$. This leads to increasingly heavy particle spectra for higher sectors --- eventually leading to sectors that are essentially pure Yang-Mills that feature strong first order phase transitions. This, of course, begs the question: at what index $i$ do said phase transitions begin?
%The actual boundary of region of interest on the Columbia plot is not precisely known and further lattice simulations are required to determine exactly where in parameter space cross-over transitions end and first order transitions begin. In lieu of such studies, we parameterize our ignorance in the form of a ``slop factor" $\kappa$. Assuming the up and down quarks of each sector to be of roughly the same scale, we determine what sector $i$ features $\kappa m_u^i \approx \Lambda_{QCD}$.
Using the methods outlined in the prior section we determine $\Lambda_{QCD}$ to have a relevant value of 
\begin{equation}\label{eqn:lambda2}
\Lambda^i_{(2)} = (m_s^i m_c^i m_b^i m_t^i)^{2/29}(\Lambda^i_{(6)})^{21/29}
\end{equation} 
at the energy scale we're interested in. $\Lambda^i_{(6)}$ is identical for all sectors and is taken to have a standard model value of $\Lambda^{(6)}_{MS} =(89 \pm 6)\, \mev$~\cite{PhysRevD.98.030001}. Rewriting Eq.~(\ref{eqn:lambda2}) in terms of standard model variables, 
\begin{equation}\label{eqn:lambda2adj}
\Lambda^i_{(2)} = (m_s m_c m_b m_t i^2)^{2/29}(\Lambda_{(6)})^{21/29}.
\end{equation}
where $m_q$ without a superscript is the mass of $q$ in the SM. We take the sector with SFOPT to be the ones when the mass of the up quark, down quark, and QCD phase transition scale are all comparable:
\begin{equation}
m^i_u \sim m_u \sqrt{i} \sim (m_s m_c m_b m_t i^2)^{2/29}(\Lambda_{(6)})^{21/29}.
\end{equation}
This can be solved for $i$:
\begin{equation}\label{eqn:critIndex}
i^c \sim \frac{(m_s m_c m_b m_t)^{4/21}(\Lambda_{(6)})^{2}}{( m_u)^{58/21}} \sim 10^6.
\end{equation}
As we will see in Sec.~\ref{sec:reheat}, in the original $N$naturalness setup~\cite{Arkani-Hamed:2016rle}, the energy dumped into the $i$th sector scales as $i^{-1}$, so there will will not be enough energy in the sectors with $i>i^c$ to see a signature of these phase transitions. 
%This critical index demonstrates that even in the most extreme cases, where first order phase transitions occur with the up quark an order of magnitude less than $\Lambda_{QCD}$, the critical index is still $\sim 1000$. Looking back to the energy density scaling in Eq.~(\ref{eqn:edSS})\textcolor{green}{refer to a particular equation} we see that, within the context of $N$naturalness, the energy density of these sectors is at most $10^{-4}$ \textcolor{green}{why not $10^{-3}$?} that of our sector. As a result, \textcolor{green}{As we will show below?} the gravitational waves generated by the standard large $i$ sectors in $N$naturalness will not generate detectable signatures. 
However, if we move away from the original $N$naturalness reheating mechanism and begin exploring mirror sectors with large vevs and with relative energy densities $\rho_i / \rho_{SM} \sim 10\%$, a possibility allowed by current constraints, we can have sectors with relatively high dark QCD scales that produce detectable gravitational waves. From Eq.~(\ref{eqn:lambda2}) we can determine the confinement scale of an arbitrary mirror sector. If we take Higgs vevs as high as the GUT scale $\sim 10^{16}$ GeV, then we can get confinement scales as high as $\sim 38 \,\gev$ \draftnote{is there a way to get this value parametrically?}.  The signals of this sector and other test cases like it are explored in Sec. \ref{sec:gw}.

@@

Ultimately, after examining the gravitational wave case produced by standard $N$naturalness, we also consider a more generic parameterization where the reheating temperature of each sector is a free parameter and is in general uncorrelated with the Higgs mass parameter. This allows us to explore a broader model space with multiple dark sectors at a huge range of scales. For these models, the reheating mechanism remains unspecified\section{Constraints}

@@

For a dark sector with one relativistic degree of freedom, its temperature must be $\sim 0.6\, T_{\rm SM}$ to not be excluded. This sector would have an energy density $\rho \sim 0.036\, \rho_{\rm SM}$ \draftnote{Explain why this isn't $.6^4$}. 

@@

\subsection{Exotic Sector Contributions}
%The original $N$naturalness paper focused on the (stronger) $N_{eff}$ constraints from the additional standard sectors and ignored the exotic sectors altogether \cite{Arkani-Hamed:2016rle}; in contrast, we take the opposite approach. Since we're more interested in a more general scenario with multiple dark phase transitions, we focus in on whether zero vev sectors can dodge these $N_{eff}$ bounds. 
We begin by computing the constraints on exotic sectors; these are significantly weaker than those for standard sectors~\cite{Arkani-Hamed:2016rle}. 
%At high scales, these sectors feature 106.75 effective degrees of freedom at high energies: 72 from quarks,  12 from charged leptons, 6 from neutrinos, 16 from gluons, 2 from photons, 6 from W bosons (no longitudinal modes due to a no symmetry breaking), 4 from the higgs sector (again, no symmetry breaking leads to 3 additional modes), for a total of 118 degress of freedom. Scaling the fermionic contributions by $7/8$ gives us 106.75 effective relativistic degrees of freedom.  After the higgses freeze out, 4 degrees of freedom are lost; when the QCD phase transition occurs 35 bosonic degrees of freedom are gained while 72 fermionic are lost, leaving 56.75 effective degrees of freedom. 
At the time of photon decoupling ($T_\gamma \sim 0.39 \,\ev$) while the temperature of the exotic sectors is lower still. This means that for sectors with small and moderate $i$, we can use Eqs.~(\ref{eqn:exoticFM}) and~(\ref{eq:ith_pion_mass}) to see that the pions will be non-relativistic  leaving at most $11.25$ effective degrees of freedom per sector from photons and neutrinos.  For very large $i$, the pions can be much lighter, but those sectors also have very little energy in them in the standard reheating scenario. 
%Since all additional sectors are colder than our own, by the time photon decoupling occurs ($T_\gamma \sim 0.39 \,\ev$), the pions of all of these sectors will, in accordance with Eq.~(\ref{eqn:gmor}), have a much higher mass than the sector temperature and thus will have frozen out long before --- leaving at most $11.25$ effective degrees of freedom per sector from photons and neutrinos. \draftnote{Pions are not decoupled at BBN time, does this in any way affect the analysis?}
Coupling the number of effective degrees of freedom per sector with the energy density scaling of $\sim 1/m_H^4$ as in Eq.~(\ref{eq:reheat_exotic}) means that the zero vev sectors have small temperature ratios. Assuming a reheating temperature of $100$ GeV and a completely uniform distribution of sectors, the temperature of the first exotic sector is slightly more than $6\%$ of our sector at reheating. Applying Eq.~(\ref{eqn:DeltaNeff}) to this particular situation gives us:
\Delta N_{eff} = \sum_i \frac{4}{7}\left(\frac{11}{4}\right)^{4/3}g_{i} \left(\frac{(T_{RH_{E1}}/T_{RH})}{i^{1/2}}\right)^4 \sim 10^{-4},
with $T_{RH_{E1}}/T_{RH}$ being the ratio of the reheat temperatures of the 1st exotic sector and our own sector ($0.06$ in standard $N$naturalness with $r = 1$).  This sum is dominated by $i=1$, the sector with the lowest Higgs mass (and thus the most energy density) gives us a contribution of $\mathcal{O} (10^{-4})$ to $\Delta N_{eff}$. Evolving the sectors thermal histories forward in time to the recombination era gives us a slightly larger value, but still of order $\mathcal{O}(10^{-4})$, well under current CMB bounds. 
It should be noted that modifying the exotic sectors' structure (e.g. adjusting the exotic sectors to have a lower higgs mass squared or clustering multiple hidden sectors close to the first exotic one) leads to a $\Delta N_{eff}$ contribution that is larger than the base $N$naturalness case. This increase is typically not excluded by current bounds, indicating a large degree of liberty in the structure and number of exotic hidden sectors.

>>>>

\subsection{Standard Sectors}
As shown in Eq.~(\ref{eqn:vevs}), for standard sectors with increasing index $i$ the vevs of said sectors increase $v_i\propto \sqrt{i}$. This leads to increasingly heavy particle spectra for higher sectors --- eventually leading to sectors that are essentially pure Yang-Mills that feature strong first order phase transitions. This, of course, begs the question: at what index $i$ do said phase transitions begin?
%The actual boundary of region of interest on the Columbia plot is not precisely known and further lattice simulations are required to determine exactly where in parameter space cross-over transitions end and first order transitions begin. In lieu of such studies, we parameterize our ignorance in the form of a ``slop factor" $\kappa$. Assuming the up and down quarks of each sector to be of roughly the same scale, we determine what sector $i$ features $\kappa m_u^i \approx \Lambda_{QCD}$.
Using the methods outlined in the prior section we determine $\Lambda_{QCD}$ to have a relevant value of 
\begin{equation}\label{eqn:lambda2}
\Lambda^i_{(2)} = (m_s^i m_c^i m_b^i m_t^i)^{2/29}(\Lambda^i_{(6)})^{21/29}
\end{equation} 
at the energy scale we're interested in. $\Lambda^i_{(6)}$ is identical for all sectors and is taken to have a standard model value of $\Lambda^{(6)}_{MS} =(89 \pm 6)\, \mev$~\cite{PhysRevD.98.030001}. Rewriting Eq.~(\ref{eqn:lambda2}) in terms of standard model variables, 
\begin{equation}\label{eqn:lambda2adj}
\Lambda^i_{(2)} = (m_s m_c m_b m_t i^2)^{2/29}(\Lambda_{(6)})^{21/29}.
\end{equation}
where $m_q$ without a superscript is the mass of $q$ in the SM. We take the sector with SFOPT to be the ones when the mass of the up quark, down quark, and QCD phase transition scale are all comparable:
\begin{equation}
m^i_u \sim m_u \sqrt{i} \sim (m_s m_c m_b m_t i^2)^{2/29}(\Lambda_{(6)})^{21/29}.
\end{equation}
This can be solved for $i$:
\begin{equation}\label{eqn:critIndex}
i^c \sim \frac{(m_s m_c m_b m_t)^{4/21}(\Lambda_{(6)})^{2}}{( m_u)^{58/21}} \sim 10^6.
\end{equation}
As we will see in Sec.~\ref{sec:reheat}, in the original $N$naturalness setup~\cite{Arkani-Hamed:2016rle}, the energy dumped into the $i$th sector scales as $i^{-1}$, so there will will not be enough energy in the sectors with $i>i^c$ to see a signature of these phase transitions. 
%This critical index demonstrates that even in the most extreme cases, where first order phase transitions occur with the up quark an order of magnitude less than $\Lambda_{QCD}$, the critical index is still $\sim 1000$. Looking back to the energy density scaling in Eq.~(\ref{eqn:edSS})\textcolor{green}{refer to a particular equation} we see that, within the context of $N$naturalness, the energy density of these sectors is at most $10^{-4}$ \textcolor{green}{why not $10^{-3}$?} that of our sector. As a result, \textcolor{green}{As we will show below?} the gravitational waves generated by the standard large $i$ sectors in $N$naturalness will not generate detectable signatures. 
However, if we move away from the original $N$naturalness reheating mechanism and begin exploring mirror sectors with large vevs and with relative energy densities $\rho_i / \rho_{SM} \sim 10\%$, a possibility allowed by current constraints, we can have sectors with relatively high dark QCD scales that produce detectable gravitational waves. From Eq.~(\ref{eqn:lambda2}) we can determine the confinement scale of an arbitrary mirror sector. If we take Higgs vevs as high as the GUT scale $\sim 10^{16}$ GeV, then we can use Eq.~(\ref{eqn:lambda2adj}) to get confinement scales as high as $\sim 38 \,\gev$.  The signals of this sector and other test cases like it are explored in Sec. \ref{sec:gw}.

@@

Ultimately, after examining the gravitational wave case produced by standard $N$naturalness, we also consider a more generic parameterization where the reheating temperature of each sector is a free parameter and is in general uncorrelated with the Higgs mass parameter. This allows us to explore a broader model space with multiple dark sectors at a huge range of scales. For these models, the reheating mechanism remains unspecifie.d\section{Constraints}

@@

For a dark sector with one relativistic degree of freedom, its temperature must be $T_{DS}\sim 0.6\, T_{\rm SM}$ to not be excluded. Applying the energy density formula \cite{Trodden:2004st},
\rho_i = \frac{\pi^2}{30}g_i T_i^4,
to both said dark sector and the SM then taking the ratio indicates that the dark sector would have an energy density $\rho \sim 0.038\, \rho_{\rm SM}$. 

@@

\subsection{Exotic Sector Contributions}
%The original $N$naturalness paper focused on the (stronger) $N_{eff}$ constraints from the additional standard sectors and ignored the exotic sectors altogether \cite{Arkani-Hamed:2016rle}; in contrast, we take the opposite approach. Since we're more interested in a more general scenario with multiple dark phase transitions, we focus in on whether zero vev sectors can dodge these $N_{eff}$ bounds. 
We begin by computing the constraints on exotic sectors; these are significantly weaker than those for standard sectors~\cite{Arkani-Hamed:2016rle}. 
%At high scales, these sectors feature 106.75 effective degrees of freedom at high energies: 72 from quarks,  12 from charged leptons, 6 from neutrinos, 16 from gluons, 2 from photons, 6 from W bosons (no longitudinal modes due to a no symmetry breaking), 4 from the higgs sector (again, no symmetry breaking leads to 3 additional modes), for a total of 118 degress of freedom. Scaling the fermionic contributions by $7/8$ gives us 106.75 effective relativistic degrees of freedom.  After the higgses freeze out, 4 degrees of freedom are lost; when the QCD phase transition occurs 35 bosonic degrees of freedom are gained while 72 fermionic are lost, leaving 56.75 effective degrees of freedom. 
At the time of photon decoupling ($T_\gamma \sim 0.39 \,\ev$) while the temperature of the exotic sectors is lower still. This means that for sectors with small and moderate $i$, we can use Eqs.~(\ref{eqn:exoticFM}) and~(\ref{eq:ith_pion_mass}) to see that the pions will be non-relativistic  leaving at most $11.25$ effective degrees of freedom per sector from photons and neutrinos.  For very large $i$, the pions can be much lighter, but those sectors also have very little energy in them in the standard reheating scenario. 
%Since all additional sectors are colder than our own, by the time photon decoupling occurs ($T_\gamma \sim 0.39 \,\ev$), the pions of all of these sectors will, in accordance with Eq.~(\ref{eqn:gmor}), have a much higher mass than the sector temperature and thus will have frozen out long before --- leaving at most $11.25$ effective degrees of freedom per sector from photons and neutrinos. \draftnote{Pions are not decoupled at BBN time, does this in any way affect the analysis?}
Coupling the number of effective degrees of freedom per sector with the energy density scaling of $\sim 1/m_H^4$ as in Eq.~(\ref{eq:reheat_exotic}) means that the zero vev sectors have small temperature ratios. Assuming a reheating temperature of $100$ GeV and a completely uniform distribution of sectors, the temperature of the first exotic sector is slightly more than $6\%$ of our sector at reheating. Applying Eq.~(\ref{eqn:DeltaNeff}) to this particular situation gives us:
\begin{equation}
\Delta N_{eff} = \sum_i \frac{4}{7}\left(\frac{11}{4}\right)^{4/3}g_{i} \left(\frac{(T_{RH_{E1}}/T_{RH})}{i^{1/2}}\right)^4 \sim 10^{-4},
\end{equation}     
with $T_{RH_{E1}}/T_{RH}$ being the ratio of the reheat temperatures of the 1st exotic sector and our own sector ($0.06$ in standard $N$naturalness with $r = 1$).  This sum is dominated by $i=1$, the sector with the lowest Higgs mass (and thus the most energy density) gives us a contribution of $\mathcal{O} (10^{-4})$ to $\Delta N_{eff}$. Evolving the sectors thermal histories forward in time to the recombination era gives us a slightly larger value, but still of order $\mathcal{O}(10^{-4})$, well under current CMB bounds. 
It should be noted that modifying the exotic sectors' structure (e.g. adjusting the exotic sectors to have a lower higgs mass squared or clustering multiple hidden sectors close to the first exotic one) leads to a $\Delta N_{eff}$ contribution that is larger than the base $N$naturalness case. This increase is typically not excluded by current bounds, indicating a large degree of liberty in the structure and number of exotic hidden sectors.

\subsection{Standard Sector Contributions}

Within the context of vanilla $N$naturalness, the majority of contributions arise from standard sectors. This is explored in a sufficient amount of depth in \cite{Arkani-Hamed:2016rle}; here we briefly summarize these arguments. All additional standard sectors are very similar to our own: they have the same particle content and couplings and differ only by the higgs mass. As our sector is taken to be the lightest so as to be preferentially reheated, every other standard sector features an earlier freeze-out of their respective particles. This ultimately leads to each sector having at most the same number of relativistic degrees of freedom as the SM.

 \cite{Arkani-Hamed:2016rle} expresses the standard sector contributions as:
\begin{equation}
\Delta N_{eff} = \frac{1}{\rho^{us}_\nu} \sum_{i\neq us}. \rho_i.
\end{equation}
In the case that the reheaton is lighter than the lightest higgs (ours), this can be expressed as
\begin{equation}
\begin{split}
\Delta N_{eff} &\sim \sum^{N_b}_{i=1}\frac{1}{2i+1}+\frac{y_c^2}{y_b^2}\sum^{N_c}_{i=N_b + 1}\frac{1}{2i+1}\\
& \simeq \frac{1}{2}\left(\log 2N_b + \frac{y_c^2}{y_b^2}\log \frac{N_c}{N_b}\right)
\end{split}
\end{equation}
with $y_{c,b}$ representing the charm and bottom yukawa couplings and 
\begin{equation}
N_{b,c} = \left(\frac{m_\phi^2}{8m^2_{b,c}}-\frac{1}{2}\right)
\end{equation}
with $m_\phi$ being the mass of the reheaton.

Application of these results indicates that for a majority of the parameter space, vanilla $N$naturalness requires fine-tuning ($r$ in Eq.~(\ref{eqn:vevs}) set to a value $< 1$). Numerical results for the fine-tuning required for various reheaton masses were presented in \cite{Arkani-Hamed:2016rle}.
